Best practices for forecasting changes in biodiversity: an example using breeding birds

### 2017-08-28 manuscript/draft.Rmd
Edits to Discussion
* Rework first 4 paragraphs back into 3 for better flow
* Combine two time-scale paragraphs and reduce emphasis on observed convergence
* Remove paragraph level headers and improve overall structure and flow instead
* Add reference for not being able to address imperfect detection
* General editorial work throughout

**Richness data.** Bird species richness was obtained from the North American
Breeding Bird Survey (BBS) [@pardieck2017] using the Data Retriever Python
package [@morris2013] and rdataretriever R package [@mcglinn2017]. The BBS data
was filtered to exclude all nocturnal, cepuscular, and aquatic species [since
these species are not well sampled by BBS methods; @hurlbert2005], as well as
unidentified species, and hybrids. All data from surveys that did not meet BBS
quality criteria were also excluded.

@@

Forecasting is an emerging imperative in ecology; as such, the field needs to
develop and follow best practices for conducting and evaluating ecological
forecasts. We have laid out ideas for a number of these practices (Box 1) and
attempted to implement them in a single study that builds and evaluates
forecasts of biodiversity in the form of species richness. The results of this
effort are both promising and humbling. When comparing forecasts across sites,
many different approaches to forecasting produce reasonable forecasts. If a site
is predicted to have a high number of species in the future, relative to other
sites, it generally does. However (after controlling for observer differences),
none of the methods we evaluated could reliably tell us whether site-level
richness would increase or decrease over time (Figure 6), which is generally the
stated purpose of these forecasts. As a result, our baseline models, which did
not attempt to anticipate changes in richness over time, generally provided the
best forecasts for future biodiversity.

These results raise a number of important issues. First, we should be skeptical
of predictions about future biodiversity from models that have not been
validated in terms of repeated observations at the same site. Since site-level
richness is relatively stable, model evaluations that show high
accuracy across spatial gradients may not indicate much about a model's ability
to predict changes over time. This issue has also recently been highlighted as a
potential downside of using SDMs to predict the future locations of individual
species [@rapacciuolo2012; @oedekoven2017]. This result is particularly sobering
because SDMs form the main foundation for estimates of the predicted loss of
biodiversity to climate change [@thomas2004; @thuiller2011; @urban2015].

Second, our results highlight the importance of comparing multiple modeling
approaches when conducting ecological forecasts, and in particular, the value of
comparing results to simple baselines to avoid over-interpreting the information
present in these forecasts (Box 1). Disciplines that have more mature
forecasting cultures often do this by reporting "forecast skill", i.e., the
improvement in the forecast relative to a simple baseline [@jolliffe2003]. We
recommend following the advice of @ye2015 and adopting this approach in future
ecological forecasting research. Our results also highlight the need for using
hindcasting and time-series data for assessing the effectiveness of forecasting
methods. This practice allows comparisons to baseline models and also
facilitates the assessment of predicted dynamics within sites or regions.

One of the main alternatives to stacked SDMs is modeling richness directly using
environmental variables and using the resulting models to make forecasts. Two
studies compared the SDM and direct richness modeling approaches, and reported
that the methods yielded equivalent results for forecasting diversity
[@algar2009; @distler2015]. While our results generally support the idea of
rough equivalence between stacked SDMs and direct richness modeling for point
estimates, they also show that stacked SDMs dramatically understimate the
uncertainty. For this reason, direct modeling of richness is a better overall
approach to forecasting richness. A similar result is seen when comparing joint
species distribution models (JSDMs) to stacked single species distribution
models. For point estimates, the joint distribution models are roughly
equivalent to stacked SDMs, but the JSDMs provide much better estimates of
uncertainty [@harris2015; @warton2015]. In fact, JSDMs and direct richness modeling
provide some of the best estimates of uncertainty across all modeling
approaches. This highlights the importance of evaluating models based on their
uncertainty as well as their mean estimates (Box 1). Not doing so is
particularly problematic in this case, because the summed species distribution
models are overly confident, which means that these models could mislead
ecologists into believing that richness would be restricted to a much narrower
range than would actually occur.

[[Note: this paragraph may no longer be accurate!]] For the error metric that
evaluates both models' point estimates and their uncertainty simultaneously, the
JSDM and the nonlinear richness model converge towards the time-series baselines
at the longest time-scales assessed. This convergence at longer time scales
highlights the importance of considering how forecasting method performance
changes with the distance into the future being predicted (Box 1). The value of
this approach, and associated forecast horizons, has been raised for knowing how
far into the future a model can be effectively used for forecasting
[@petchey2015], but our results suggest a broader value to this approach for
considering the potential importance of different models and processes for
making forecasts at different scales. Our results show that for forecasts 1-3
years in the future, baselines and time-series based approaches outperform all
other methods using integrative metrics. At these time scales, environmental
changes are relatively small and changes in the biota may lag behind changes in
the environment. However, as the timescale of the forecast increases to a decade
the JSDMs and environmental richness models get closer to the accuracy of the
time-series methods and even surpass some of those methods. As the forecast
distance increases the amount of environmental change is expected to increase
and the system will have more time to respond, potentially leading to shifts
towards improved relative performance of the models incorporating environmental
and ecological information. If this shift with forecast distance in the relative
importance of different processes and models continues, this could lead to JSDMs
and environmental richness models outperforming time-series based approaches at
sufficiently long time scales. These results suggest that assessment of how
forecast performance changes with time lag is not only important for determining
how far in the future to forecast, but also for determining which models and
processes are most relevant for making forecasts at different time scales. For
example ensemble forecasts could be created which incorporate these time lags
into the weights for individual models.

It is also possible that models that include species' relationships to their
environments or direct environmental constraints on richness will continue to
produce forecasts that are worse than simply assuming the systems are static.
This would be expected to occur if richness in these systems is not changing
over the relevant time scales (e.g. 10-30 years in this study), which would make
simpler models with no directional change more appropriate. Recent suggestions
that local scale richness in some systems is not changing directionally at
multi-decadal scales supports this possibility [@brown2001; @ernest2001;
@dornelas2014; @vellend2013]. This lack of change in richness may be expected
even in the presence of substantial changes in environmental conditions and
species composition at a site due to the replacement of species from the
regional pool [@brown2001; @ernest2001]. On average, the Breeding Bird Survey
sites used in this study show little change in richness [site-level SD of `r
numbers$resid_sd` species, after controlling for differences among observers;
see also @lasorte2005]. The absence of rapid change in this dataset is
beneficial for the absolute accuracy of forecasts across different sites: when a
past year's richness is already known, it is easy to estimate a future richness.
The site-level stability of the BBS data also explains why stacked SDMs perform
relatively well at predicting future richness, despite failing to capture
changes in richness over time. However, this stability also makes it difficult
to improve forecasts relative to simple baselines, since those baselines are
already close to representing what is actually occurring in the system. These
results suggest that single-site models should be actively considered for
forecasts of richness and other stable aspects of biodiversity. Our results also
suggest that future efforts to understand and forecast biodiversity should also
focus on species composition, since lower-level processes are expected to be
more dynamic [@ernest2001; @dornelas2014] and contain more useful information
[@harris2015].

**Observer effects.** In addition to consideration of the different process
models used for forecasting it is important to consider the observation models.
When working with any ecological dataset, there are imperfections in the
sampling process that have the potential to influence results. With large scale
surveys and citizen science datasets, such as the Breeding Bird Survey, these
issues are potentially magnified by the large number of different observers and
by major differences in the habitats and species being surveyed [@sauer1994].
Accounting for differences in observers reduced the average error in our point
estimates and also improved the coverage of the confidence intervals. In
addition, controlling for observer effects resulted in changes in which models
performed best, most notably reducing the relative performance of the naive
model's point estimates. This suggests that, prior to accounting for observer
effects, the naive model performed well because it was capable of accommodating
rapid shifts in estimated richness introduced by changes in the observer. These
kinds of rapid changes were difficult for the other single-site models to
accommodate, and so the average performance of the ARIMA and average models
improved once this source of observation error was addressed. This demonstrates
that modeling observation error can be important for properly estimating and
reducing uncertainty in forecasts and can also lead to changes in the best
methods for forecasting. We did not address differences in detection probability
across species and sites [@boulinier1998] since there is no clear way to address
this issue without making strong assumptions about the data (i.e., assuming
there is no biological variation in stops along a route), but this would be a
valuable addition to future forecasting models.

**Environmental uncertainty.** Future biodiversity forecasting efforts will also
need to begin to address the additional uncertainty introduced by the error
inherent in the process of forecasting the environmental conditions that
ecologists use as predictor variables. In this, and other hindcasting studies,
the environmental conditions for the "future" are known because the data has
already been observed. However, in real forecasts the environmental conditions
themselves have to be predicted, and environmental forecasts will also have
uncertainty and bias. Ultimately, ecological forecasts that use environmental
data will therefore be more uncertain than our current hindcasting efforts.
Thus, it is important to correctly incorporate this uncertainty in the predictor
variables into forecasting models [@clark2001; @dietze2017], as we did with the
uncertainty in observer-level effects. Difficulty in forecasting future
environmental conditions at small scales will present continued challenges for
models incorporating environmental variables, and this may result in a continued
advantage for simple single-site approaches.

**Multi-timescale forecasts.** Another challenge for ecological forecasting is
the timescales of typical forecasts. We evaluated forecasts using yearly
timesteps up to a decade into the future. In contrast, many ecological forecasts
are temporally aggregated to 5-30 year timesteps and projected up to a century
into the future. These are commonly beyond the career or even lifespan of the
researchers, which makes it impossible to directly validate the models'
predictions. Currently, assessments at these long time-scales can only be made
with a small number of opportune datasets. For example, hindcasts have been used
to assess models of species richness [@algar2009; @distler2015] and distribution
[@rapacciuolo2012; @moran-ordonez2017; @araujo2005; @eskildsen2013] using data
aggregated across similar time scales. More evaluation
studies like these are needed but there is a paucity of adequate long term data
available. Short term forecasts will allow for relatively rapid validation of
models by researchers without needing to wait for enough data to accumulate
[@tredennick2016; @dietze2016]. This does not mean that they should replace
large scale long-term forecasts. As discussed above, drivers of species richness
are likely to differ at different temporal scales [@rosenzweig1995; @white2004;
@white2007]. Thus, short-term forecasts should be expected to add value to
scientists and policymakers, and inform (not replace) long-term forecasts.
Hopefully, by evaluating trends in the shorter term forecasts as done here it
will be possible to also use the information from short-term forecasting to
improve longer-term estimates.

**Conclusions.** The science of forecasting biodiversity, and ecology more
broadly, remains in its infancy and it is important to consider the general
inability of forecasting methods to improve on simple baselines in that context.
When weather forecasting first started the forecasts were likewise worse than
simple baselines [@mcgill2012]. One practice that helped weather forecasts to
improve was publicly making a large number of predictions, which allowed
different forecasting approaches to be regularly assessed and refined
[@mcgill2012; @silver2012]. This suggests that it is important for ecologists to
start regularly making and evaluating ecological forecasts on real data, even if
they perform poorly, and to make these forecasts publicly available for
assessment. These forecasts should include both short-term predictions, which
can be assessed quickly, and mid- to long-term forecasts, which can help
ecologists to assess long time-scale processes and determine how far into the
future we can successfully forecast [@tredennick2016; @dietze2017]. Forecasts
based on the models in this paper (from now until 2050) are openly archived on
Zenodo, so that we and others can assess how well they perform. We also plan to
evaluate these forecasts and report the results as each new year of BBS data
becomes available at the same URL [@zenodo]. Weather forecasting has continually
improved throughout its history [@bauer2015], in part due to making and
evaluating public forecasts [@mcgill2012], and we hope this practice will help
ecology do the same.

Making successful ecological forecasts will be challenging. Ecological systems
are complex, fundamental theory is less refined than simpler physical/chemical
systems, and we currently lack the large amounts of data that produce effective
forecasts through machine learning. Despite this, we believe that progress can be
made if we build an active forecasting culture in ecology that builds and
assesses forecasts in ways that will allow us to improve the effectiveness of
ecological forecasts most rapidly (Box 1). This includes expanding the scope of
the ecological and environmental data we work with, paying attention to
uncertainty in both model building and forecast evaluation, and rigorously
assessing forecasts using a combination of hindcasting, archived forecasts, and
comparisons to simple baselines.

>>>>

**Richness data.** Bird species richness was obtained from the North American
Breeding Bird Survey (BBS) [@pardieck2017] using the Data Retriever Python
package [@morris2013] and rdataretriever R package [@rdataretriever]. The BBS
data was filtered to exclude all nocturnal, cepuscular, and aquatic species
[since these species are not well sampled by BBS methods; @hurlbert2005], as
well as unidentified species, and hybrids. All data from surveys that did not
meet BBS quality criteria were also excluded.

@@

Forecasting is an emerging imperative in ecology; as such, the field needs to
develop and follow best practices for conducting and evaluating ecological
forecasts [@clark2001]. We have used a number of these practices (Box 1) in a
single study that builds and evaluates forecasts of biodiversity in the form of
species richness. The results of this effort are both promising and
humbling. When comparing forecasts across sites, many different approaches to
forecasting produce reasonable forecasts (Figure 3). If a site is predicted to
have a high number of species in the future, relative to other sites, it
generally does. However, none of the methods evaluated reliably determined how
site-level richness changes over time (Figure 6), which is generally the stated
purpose of these forecasts. As a result, baseline models, which did not attempt
to anticipate changes in richness over time, generally provided the best
forecasts for future biodiversity. It should be noted that the scope of
inference for our results is limited to local scale forecasts for breeding birds
in North America. As such we cannot draw meaningful conclusions about global
scale forecasts of biodiversity loss or extinction rates, or about about
forecasting biodiversity changes in areas of the world like the tropics that are
expected to be heavily impacted by global change.

The most commonly used methods for forecasting future biodiversity, SDMs and
macroecological models, produced worse forecasts than time-series models and
simple baselines. This highlights the need for using hindcasting and time-series
data for assessing the effectiveness of forecasting methods and suggests that
predictions about future biodiversity change that have not been validated in
this way should be viewed with skepticism. Since site-level richness is
relatively stable, model evaluations that show high accuracy across spatial
gradients may not indicate that a model is able to predict changes over
time. This issue has also recently been highlighted as a potential downside of
using SDMs to predict the future locations of individual species
[@rapacciuolo2012; @oedekoven2017]. This result is particularly sobering because
SDMs are one of the main foundations for estimates of the predicted loss of
biodiversity to climate change [@thomas2004; @thuiller2011; @urban2015]. These
results also highlight the importance of comparing multiple modeling approaches
when conducting ecological forecasts, and in particular, the value of comparing
results to simple baselines to avoid over-interpreting the information present
in these forecasts [Box 1]. Disciplines that have more mature forecasting
cultures often do this by reporting "forecast skill", i.e., the improvement in
the forecast relative to a simple baseline [@jolliffe2003]. We recommend
following the example of @ye2015 and adopting this approach in future ecological
forecasting research.

When comparing different methods for forecasting our results demonstrate the
importance of considering uncertainty [Box 1; @clark2001; @dietz2016]. Previous
comparisons between stacked SDMs and macroecological models reported that the
methods yielded equivalent results for forecasting diversity [@algar2009;
@distler2015]. While our results support this equivalence for point estimates,
they also show that stacked SDMs dramatically underestimate uncertainty. For
this reason, direct modeling of richness is a better overall approach to
forecasting richness in this case. A similar result is seen when comparing joint
species distribution models (JSDMs) to stacked single species distribution
models. For point estimates, the joint distribution models are roughly
equivalent to stacked SDMs, but the JSDMs provide better estimates of
uncertainty [@harris2015; @warton2015]. In fact, JSDMs and direct richness
modeling provide some of the best estimates of uncertainty across all modeling
approaches. Failing to consider uncertainty in model comparisons can result
changes in the choice of methods and can lead to over confident models, which
could result in predictions that biodiversity would be restricted to a narrower
range than would actually occur.

The accuracy for forecasts is expected to decline as forecasts are made further
into the future, so it is important to evaluate forecasts at different lead
times [Box 1; @petchey2015]. For deviance (the error metric that evaluates both
models' point estimates and their uncertainty simultaneously), there is some
indication that the JSDM and the nonlinear richness model converge towards the
time-series and baseline models as the lead time of the forecast increases. This
potential convergence is biologically plausible because at short lead times,
environmental changes are relatively small and changes in the biota may lag
behind changes in the environment. However, as the lead time of the forecast
increases the amount of environmental change should increase and the systems
will have more time to respond, potentially resulting in shifts towards improved
relative performance of models incorporating environmental and ecological
information. This is potentially important for interpreting our results because
we only evaluated annual forecasts up to a decade into the future. In contrast,
many ecological forecasts are temporally aggregated to 5-30 year timesteps and
have lead times of up to a century. These long-term forecasts are difficult to
evaluate given the small number of century scale datasets, but are important for
understanding changes in biodiversity at some of the lead times relevant for
conservation and management. Two studies have assessed models of species
richness at longer lead times [@algar2009; @distler2015], but the results were
not compared to baseline or time-series models (in part due to data limitations)
making them difficult to compare to our results. Given the potential changes
with lead time of the models and processes are most relevant to forecasting, it
is important that more long lead time evaluation studies are conducted and that
the inferences drawn from short-term and long-term forecasting studies be
compared. For true forecasts, short term forecasts will allow for relatively
rapid validation of models by researchers without needing to wait decades for
data to accumulate [@tredennick2016; @dietze2016]. However, since drivers of
species richness differ at different temporal scales [@rosenzweig1995;
@white2004; @white2007], short-term forecasts should be viewed as informing, not
replacing, long-term forecasts [@tredennick2016].

While it is possible that models that include species' relationships to their
environments or direct environmental constraints on richness will provide better
fits at longer lead times, it is also possible that they will continue to
produce forecasts that are worse than baselines assuming that the systems are
static.  This would be expected to occur if richness in these systems is not
changing over the relevant multi-decadal time scales, which would make simpler
models with no directional change more appropriate. Recent suggestions that
local scale richness in some systems is not changing directionally at
multi-decadal scales supports this possibility [@brown2001; @ernest2001;
@vellend2013; @dornelas2014]. A lack of change in richness may be expected even
in the presence of substantial changes in environmental conditions and species
composition at a site due to replacement of species from the regional pool
[@brown2001; @ernest2001]. On average, the Breeding Bird Survey sites used in
this study show little change in richness [site-level SD of `r numbers$resid_sd`
species, after controlling for differences among observers; see also
@lasorte2005]. The absence of rapid change in this dataset is beneficial for the
absolute accuracy of forecasts across different sites: when a past year's
richness is already known, it is easy to estimate future richness.  The
site-level stability of the BBS data also explains why SDMs and macroecological
models perform relatively well at predicting future richness, despite failing to
capture changes in richness over time. However, this stability also makes it
difficult to improve forecasts relative to simple baselines, since those
baselines are already close to representing what is actually occurring in the
system. These results suggest that single-site models should be actively
considered for forecasts of richness and other stable aspects of
biodiversity. Our results also suggest that future efforts to understand and
forecast biodiversity should incorporate species composition, since lower-level
processes are expected to be more dynamic [@ernest2001; @dornelas2014] and
contain more useful information [@harris2015].

Future biodiversity forecasting efforts also need to address the uncertainty
introduced by the error in forecasting the environmental conditions that are
used as predictor variables. In this, and other hindcasting studies, the
environmental conditions for the "future" are known because the data has already
been observed. However, in real forecasts the environmental conditions
themselves have to be predicted, and environmental forecasts will also have
uncertainty and bias. Ultimately, ecological forecasts that use environmental
data will therefore be more uncertain than our current hindcasting efforts. It
is important to correctly incorporate this uncertainty in the predictor
variables into forecasting models [@clark2001; @dietze2017]. Limitations in
forecasting future environmental conditions, particularly at small scales, will
present continued challenges for models incorporating environmental variables,
and this may result in a continued advantage for simple single-site approaches.

In addition to comparing and improving the process models used for forecasting
it is important to consider the observation models.  When working with any
ecological dataset, there are imperfections in the sampling process that have
the potential to influence results. With large scale surveys and citizen science
datasets, such as the Breeding Bird Survey, these issues are potentially
magnified by the large number of different observers and by major differences in
the habitats and species being surveyed [@sauer1994]. Accounting for differences
in observers reduced the average error in our point estimates and also improved
the coverage of the confidence intervals. In addition, controlling for observer
effects resulted in changes in which models performed best, most notably
reducing the relative performance of the naive model's point estimates. This
suggests that, prior to accounting for observer effects, the naive model
performed well because it was capable of accommodating rapid shifts in estimated
richness introduced by changes in the observer. These kinds of rapid changes
were difficult for the other single-site models to accommodate, and so the
average performance of the ARIMA and average models improved once this source of
observation error was addressed. This demonstrates that modeling observation
error can be important for properly estimating and reducing uncertainty in
forecasts and can also lead to changes in the best methods for forecasting [Box
1]. Another key aspect of an ideal observation model is imperfect detection. In
this study, we did not address differences in detection probability across
species and sites [@boulinier1998] since there is no clear way to address this
issue using NA Breeding Bird Survey data without making strong assumptions about
the data [i.e., assuming there is no biological variation in stops along a
route; @white2010], but this would be a valuable addition to future forecasting
models.

The science of forecasting biodiversity remains in its infancy and it is
important to consider weaknesses in current forecasting methods in that
context. In the beginning weather forecasts were also worse than simple
baselines, but these forecasts have continually improved throughout the history
of the field [@mcgill2012; @silver2012; @bauer2015]. One practice that lead to
improvements in weather forecasts was that large numbers of forecasts were made
publicly, allowing different approaches to be regularly assessed and refined
[@mcgill2012; @silver2012]. To facilitate this kind of improvement, it is
important for ecologists to start regularly making and evaluating real
ecological forecasts, even if they perform poorly, and to make these forecasts
openly available for assessment [@mcgill2012; @dietze2016]. These forecasts
should include both short-term predictions, which can be assessed quickly, and
mid- to long-term forecasts, which can help ecologists to assess long time-scale
processes and determine how far into the future we can successfully forecast
[@dietze2016; @tredennick2016]. Forecasts out to 2050 based on the models in
this paper have been openly archived [@zenodo], so that we and others can assess
how well they perform. We plan to evaluate these forecasts and report the
results as each new year of BBS data becomes available, and make iterative
improvements to the forecasting models in response to these assessments.

Making successful ecological forecasts will be challenging. Ecological systems
are complex, our fundamental theory is less refined than for simpler physical
and chemical systems, and we currently lack the scale of data that often
produces effective forecasts through machine learning. Despite this, we believe
that progress can be made if we develop an active forecasting culture in ecology
that builds and assesses forecasts in ways that will allow us to improve the
effectiveness of ecological forecasts more rapidly [Box 1; @mcgill2012;
@dietze2016]. This includes expanding the scope of the ecological and
environmental data we work with, paying attention to uncertainty in both model
building and forecast evaluation, and rigorously assessing forecasts using a
combination of hindcasting, archived forecasts, and comparisons to simple
baselines.
